Embedding Intuitionistic into Classical Logic

### 2022-07-15 CSL/lipics-v2021-EIICL.tex
pass over overview finished

We shall now give an overview of the main results and highlight the key arguments. Recall that our goal is to give an effective translation procedure that, for a given formula $\varphi$, yields a formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. We start with the predicate case. The key arguments are the same as in the first-order case while it is technically simpler and less cluttered and therefore more clear.

Before performing the main transformation we propose a preprocessing to a simpler form. A popular pre-processing step in classical automated reasoning is the Tseytin transformation~\cite{tseitin1983complexity}. It gives an equisatisfiable sentence in conjunctive normal form. Eliminating all implications is not possible in intuitionistic logic since $A\to B$ is not equivalent to $\neg A\vee B$. Still we propose a similar transformation. A nice feature will be that the non-classical content will be encapsulated in formulas of type $(A\to B)\to C$, i.e. if there are no such formulas in the transformed formula then the classical validity of the original formula immediately implies intuitionistic validity.

\begin{theorem}
	For every propositional formula $\varphi$ there exists an effective procedure that yields an atom $P$ and a set of $\mathcal S$ of formulas  of one of the forms
	$$A\to (B\wedge C), (A\wedge B)\to C, A\to (B\vee C), (A\vee B)\to C, (A\to B)\to C$$for atomic $A, B, C$ such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The size of $\mathcal S$ and time complexity of the transformation are linear in the input size.
\end{theorem}

The main transformation then proceeds in three steps.

First we encode as a first-order sentence that the considered formula holds in every Kripke frame. Since Kripke frames for propositional logic over a fixed alphabet of propostional variables form a first-order theory this is rather straightforward.

Next we eliminate some quantifiers via Herbrandization. The only quantifiers eliminated occur in formulas derived from formulas of type $(A\to B)\to C$.

This leaves us with a formula with only weak $\forall$-quantifiers. We can then argue that if there is a counter-model for the resulting formula then there is a counter-model with a certain Kripke frame completely determined by the number of formulas of type $(A\to B)\to C$. This allows us to eliminate the remaining quantifiers by simply enumerating the worlds in that Kripke model.

All in all this leaves us with the following:

\begin{theorem}
	Let $\mathcal S$ be as above. Let $\mathcal F_\to\subseteq\mathcal S$ denote formulas $\psi$ of the form $(A_\psi\to B_\psi)\to C_\psi$ and $\Lambda$ denote the set of sequences without repetition over $\mathcal F_\to$. For each atom $A$ and $k\in\Lambda$ consider a new atom $A^k$. Obtain $\mathcal S^\#$ by including formulas as follows
	\begin{itemize}
		\item $A^k\to A^{k\psi}$ for each atom $A$ occurring in $\mathcal S$, $k\in\Lambda$ and $\psi\in\mathcal F_\to$ not occurring in $k$.
		\item $A^k\to (B^k\circ C^k)$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
		\item $(A^k\circ B^k)\to C^k$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
		\item $(A^{k\psi}\to B^{k\psi})\to C^k$ for $\psi = (A\to B)\to C\in\mathcal S$, $k\in\Lambda$ if $\psi$ does not occurr in $k$.
	\end{itemize}
	Then $\bigwedge S\to P$ is intuitionistically valid iff $\bigwedge S^\#\to P^\epsilon$ is classically valid, where $\epsilon$ denotes the empty sequence. The size of $\mathcal S^\#$ is in $\mathcal O(|\mathcal S|\cdot2^{|\mathcal F_\to|\cdot\log(|\mathcal F_\to|)})$. Futhermore there is an effective procedure for translating counter-models between the sentences.
\end{theorem}

Examining the proof of the above theorem we will be able to notice that due to its tree form the constructed Kripke frame lends itself as an interpretation for a certain quantified boolean formula. We can show that this QBF is satisfiable if and only if there exists a counter-model for our original formula.

@@

We then move to the first-order case. Here things are a bit more involved but the underlying approach is the same. First we again perform a Tseytin-like transformation. Here the non-classical content is encapsulated by formulas of type $(A\to B)\to C$ and $\forall xA\to B$.

\begin{theorem}
	For every predicate formula $\varphi$ there exists an effective procedure that yields an atom $P$ and a set $\mathcal S$ of formulas of the forms
	$$\forall \vec z(A(\vec a)\to (B(\vec b)\wedge C(\vec c))), \forall \vec z((A(\vec a)\wedge B(\vec b))\to C(\vec c)),$$$$\forall \vec z(A(\vec a)\to (B(\vec b)\vee C(\vec c))),
	\forall \vec z((A(\vec a)\vee B(\vec b))\to C(\vec c)),$$$$ \forall \vec z((A(\vec a)\to B(\vec b))\to C(\vec c)),\forall \vec z(\forall xA(\vec a)\to B(\vec b)),$$$$ \forall \vec z(A(\vec a)\to\forall xB(b)), \forall \vec z(\exists xA(\vec a)\to B(\vec b)), \forall \vec z(A(\vec a)\to\exists xB(b))$$for atomic $A, B, C$ and vectors of variables $\vec z, \vec a, \vec b, \vec c$ such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The size of $\mathcal S$ and time complexity of the transformation are linear in the input size.
\end{theorem}

We then proceed quite similarly to the propositional case. Note however that one difficulty arises from the fact that our (classical) domain will now contain on the one hand worlds in the Kripke frame but on the other hand also proper domain elements. We resolve this apparent conflict by introducing a special binary predicate $E$, inspired by~\cite{iemhoff2010eskolemization}, encoding which domain elements exists at which world. In particular $E(x, u)$ entails that $x$ represents a domain element and $u$ a world.

The Herbrandization is completely analogous the the propositional case.

The real difference arises in the final step: While we can also establish a universal Kripke Frame that is parametrized only by the formulas of type $(A\to B)\to C$ and $\forall xA\to B$, it is possibly countably infinite. This is not surprising since certain intuitionistically invalid formulas like the double negation shift $\forall x\neg\neg A(x)\to \neg\neg\forall x A(x)$ don't have finite counter-models. Therefore we are not able to eliminate the weak $\forall$-quantifiers associated with the Kripke semantics. However since we started in first-order logic to begin with this is perhaps not a big deal.

\begin{theorem}
	There exists a linear-time procedure that gives for every first-order formula $\varphi$ a formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. The size of $\varphi^\#$ is in linear in the size of $|\varphi|$, however for each $n$-ary relation symbol in $\varphi$ there is a corresponding $n+1$-ary relation symbol in $\varphi^\#$ and furthermore $\varphi$ contains a new binary predicate $E$ as well as a number of new function symbols. Furthermore there is an effective translation between intuitionistic counter-models of $\varphi$ and classical counter-models of $\varphi^\#$.
\end{theorem}
The complete translation is stated in Theorem~\ref{fullFOtranslation}.

@@

\section{Equivalid normalization}

>>>>

We shall now give an overview of our main results and highlight the key arguments. Recall that our goal is to give an effective translation procedure that, for a given formula $\varphi$, yields a formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. We start with the propositional case. The key arguments are the same as in the first-order case, while it is technically simpler and less cluttered.

Before our main transformation we employ a preprocessing step akin to the Tseytin transformation~\cite{tseitin1983complexity}, which is a popular pre-processing step in classical automated reasoning: 
It gives an equisatisfiable sentence (over an extended set of propositions) in conjunctive normal form.
Eliminating all implications, however, is not possible in intuitionistic logic since $A\to B$ is not equivalent to $\neg A\vee B$. 
Still we propose a similar transformation. 
A notable feature of our transformation is that all non-classical content is encapsulated in formulas of type $(A\to B)\to C$, i.e. if there are no such formulas in the transformed formula then the classical validity of the transformed formula immediately implies intuitionistic validity.

\begin{theorem}
For every propositional formula $\varphi$ there effectively are an atom $P$ and a set of $\mathcal S$ of formulas (over an extended set of propositions) of one of the forms\\
$\text{ \hspace{1cm}} A\to (B\wedge C), (A\wedge B)\to C, A\to (B\vee C), (A\vee B)\to C, (A\to B)\to C,$\\
for atomic $A, B, C$, such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The time complexity of the transformation is linear in the input size.
\end{theorem}

The main transformation then proceeds in three steps:
%
1) We encode as a first-order sentence that the considered formula holds for every Kripke frame. 
Since Kripke frames for propositional logic over a fixed alphabet of propositional variables form a first-order theory this step is rather straightforward.
%
2) Next we eliminate some quantifiers via Herbrandization. The only quantifiers eliminated occur in formulas derived from formulas of type $(A\to B)\to C$.
%
3) We can then argue that if there is a counter-model for the resulting formula then there is a counter-model with a certain Kripke frame of bounded size completely determined by the number of formulas of type $(A\to B)\to C$.
This allows us to eliminate the remaining quantifiers by simply enumerating the worlds in that Kripke model.

Summarizing, we obtain the following result:

\begin{theorem}
\label{thm:reduction-propositional}
	Let $\mathcal S$ be as above. Let $\mathcal F_\to\subseteq\mathcal S$ denote formulas $\psi$ of the form $(A\to B)\to C$ and $\Lambda$ denote the set of sequences without repetition over $\mathcal F_\to$. For each atom $A$ and $k\in\Lambda$ consider a new atom $A^k$. Obtain $\mathcal S^\#$ by including the following formulas:
	\begin{itemize}
		\item $A^k\to A^{k\psi}$ for each atom $A$ occurring in $\mathcal S$, $k\in\Lambda$ and $\psi\in\mathcal F_\to$ not occurring in $k$.
		\item $A^k\to (B^k\circ C^k)$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
		\item $(A^k\circ B^k)\to C^k$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
		\item $(A^{k\psi}\to B^{k\psi})\to C^k$ for $\psi = (A\to B)\to C\in\mathcal S$, $k\in\Lambda$ if $\psi$ does not occurr in $k$.
	\end{itemize}
Then, $\bigwedge S\to P$ is intuitionistically valid iff $\bigwedge S^\#\to P^\epsilon$ is classically valid, where $\epsilon$ denotes the empty sequence. The size of $\mathcal S^\#$ is in $\mathcal O(|\mathcal S|\cdot2^{|\mathcal F_\to|\cdot\log(|\mathcal F_\to|)})$. Futhermore, there is an effective procedure for translating counter-models between the sentences.
\end{theorem}

Instead of explicitly enumerating the worlds of a counter-example of bounded size (as in the reduction stated in the above theorem), we can also do this enumeration implicitly with a quantified boolean formula (QBF).
We can show that this QBF is satisfiable if and only if there exists a counter-model for our original formula:

@@

%\begin{theorem}
%	For every predicate formula $\varphi$ there exists an effective procedure that yields an atom $P$ and a set $\mathcal S$ of formulas of the forms
%	$$\forall \vec z(A(\vec a)\to (B(\vec b)\wedge C(\vec c))), \forall \vec z((A(\vec a)\wedge B(\vec b))\to C(\vec c)),$$$$\forall \vec z(A(\vec a)\to (B(\vec b)\vee C(\vec c))),
%	\forall \vec z((A(\vec a)\vee B(\vec b))\to C(\vec c)),$$$$ \forall \vec z((A(\vec a)\to B(\vec b))\to C(\vec c)),\forall \vec z(\forall xA(\vec a)\to B(\vec b)),$$$$ \forall \vec z(A(\vec a)\to\forall xB(b)), \forall \vec z(\exists xA(\vec a)\to B(\vec b)), \forall \vec z(A(\vec a)\to\exists xB(b))$$for atomic $A, B, C$ and vectors of variables $\vec z, \vec a, \vec b, \vec c$ such that $\varphi$ is intuitionistically valid if and only if $\bigwedge\mathcal S\to P$ is intuitionistically valid. The size of $\mathcal S$ and time complexity of the transformation are linear in the input size.
%\end{theorem}

We then move to the first-order case. Here the construction is a bit more involved but the underlying approach is the same.
Again we first perform a Tseytin-like transformation.
Here the non-classical content is encapsulated by formulas of type $(A\to B)\to C$ and $\forall xA\to B$.
The details of this translation are given in Section~\ref{section:tseytin}.
We then proceed similarly to the propositional case. 
Note however that one difficulty arises from the fact that our (classical) domain will now contain on the one hand worlds in the Kripke frame but on the other hand also proper domain elements. 
We resolve this apparent conflict by introducing a special binary predicate $E$, inspired by~\cite{iemhoff2010eskolemization}, encoding which domain elements exists at which world. 
In particular $E(x, u)$ entails that $x$ represents a domain element and $u$ a world.
The main transformation then again proceeds in three steps:
Step 1) and 2) are analogous to the propositional case.
The real difference arises in step 3): 
While we can also establish the existence of a canonic counter-model whose size only depends on formulas of type $(A\to B)\to C$ and $\forall xA\to B$, the size of this counter-model is countably infinite in general.
This is not surprising since certain intuitionistically invalid formulas like the double negation shift $\forall x\neg\neg A(x)\to \neg\neg\forall x A(x)$ don't have finite counter-models. 
Therefore we are not able to eliminate the $\forall$-quantifiers associated with the Kripke semantics. 
However since our translation targets first-order logic, we believe that the introduction of these quantifiers is acceptable. 
%
Summarizing, we obtain the following result:

\begin{theorem}
\label{thm:reduction-first-order-short}
There exists a linear-time procedure that gives for every first-order formula $\varphi$ a formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. The size of $\varphi^\#$ is in linear in the size of $|\varphi|$, however for each $n$-ary relation symbol in $\varphi$ there is a corresponding $n+1$-ary relation symbol in $\varphi^\#$ and furthermore $\varphi$ contains a new binary predicate $E$ as well as a number of new function symbols. Furthermore there is an effective translation between intuitionistic counter-models of $\varphi$ and classical counter-models of $\varphi^\#$.
\end{theorem}
A more detailed account of Theorem~\ref{thm:reduction-first-order-short} is given in Theorem~\ref{fullFOtranslation}.

@@

\section{Equivalid normalization}
\label{section:tseytin}

\todo{find better name for section, as the normalization works for equisatsifiability as well as for equivalidity}
